Last glacial cycle glacier erosion potential in the Alps

### 2021-01-25 articles/erosion/alps-erosion.tex
Rework the introduction, add note for Ian.

% custom commands
\newcommand{\idea}[1]{\textcolor{darkgreen}{\emph{[\textbf{IDEA:} #1]}}}
\newcommand{\note}[1]{\textcolor{darkblue}{\emph{[\textbf{NOTE:} #1]}}}
\newcommand{\todo}[1]{\textcolor{darkred}{\emph{[\textbf{TODO:} #1]}}}
\newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}

@@

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, and scientists for centuries. Indeed, its cultural
    impact is so far-reaching that in English, a non-Alpine language, the
    adjective ``alpine'' with non-capital ``a'' is commonly used to describe
    an Alpine-like, glacially modified, mountain landscape outside the Alps,
    while the proper noun ``Alps'' has been reused to nick-name glacier eroded
    mountain ranges of New Zealand, Japan, and others.

    Some mountain ranges are predominantly characterised by cirque glaciation
    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
    large-scale overdeepenings (e.g. Patagonia). But other regions, including
    the Alps, present a higher variety glacial erosional landforms, thereby
    bearing a different history of glacial erosion (Fig.~\ref{fig:landscape}).
    [The degree of glacial modification of a landscape has sometimes been used
    as a proxy for the duration of past glacier cover \aref, including for
    instance in geomorphological mapping \aref.] This paper challenges this
    assumption.

    While it has long been understood that glaciers leave a specific erosional
    imprint on the landscape, observing and quantifying the long-term erosion
    and sedimentation processes has been a challenge, not least due to the
    inaccessibility of glacier beds and the slow and stochastic nature of
    glacial erosion processes. More recent studies have moved away from process
    understanding in attempts to determine empirical erosion laws for an entire
    glacier or set of glaciers. \citet{Koppes.etal.2015} quantified sediment
    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
    predominant control of surface air temperature on rapid glacier erosion and
    calibrating a near-square-law from glacier velocity to erosion rate.
    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
    in the outlet stream of Franz-Josef Glacier, mapped their chemical
    composition to geologic zones of distinctive glacier speed, and also
    concluded to a nearly-square relationship between from basal sliding to
    erosion rate, albeit much higher erosion rates. On the other hand,
    \citet{Cook.etal.2020} assembled a global compilation of erosion rate for
    38~glaciers, showing a predominant role of glacier sliding velocity over
    climate variables, yet concluding at a sub-linear erosion law.

    These studies show that, while glacier and ice sheet velocities vary by
    several orders of magnitude, their relationship to erosion can be
    non-linear, yielding an even larger spectrum of variation for glacier
    erosion rates. Besides, cold-based glaciers have been shown to preserve
    landscape, but it remains unclear how sharp the erosional transition to
    cold-base ice is, and why some glaciated regions show traces of cold-base
    glaciations and others not.

    Here, a non-linear erosion law is applied to previously published model
    results and the patterns of modelled erosion rate and cumulative
    last glacial cycle erosion potential are analysed. The resulting
    uncertainties are expectantly enormous, yet the results are hoped to
    provide insights on the diversity of the Alpine glacial erosion landscape.

    \note{OK, maybe I need to do a literature review...}

>>>>

% custom commands
\newcommand{\idea}[1]{\textcolor{darkgreen}{\emph{[\textbf{IDEA:} #1]}}}
\newcommand{\note}[1]{\textcolor{darkblue}{\emph{[\textbf{NOTE:} #1]}}}
\newcommand{\todo}[1]{\textcolor{darkred}{\emph{[\textbf{TODO:} #1]}}}
\newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}
\newcommand{\ian}[1]{\textcolor{darkbrown}{\emph{[\textbf{IAN:} #1]}}}

@@

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, mountaineers and scientists for centuries. Its cultural
    impact is indeed so far-reaching, that in English, a non-Alpine language,
    the adjective ``alpine'' with non-capital ``a'' is now casually used to
    describe an Alpine-like, glacially modified mountain landscape outside the
    Alps, while the proper noun ``Alps'' has been applied to nick-name
    Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
    New Zealand (Southern Alps), Japan (\emph{Nihon Arupusu}), and elsewhere.

    Some mountain ranges are predominantly characterised by cirque glaciation
    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
    large-scale overdeepenings (e.g. Patagonia). But other regions, including
    the Alps, present a higher variety glacial erosional landforms
    (Fig.~\ref{fig:landscape}), whose implications on glacial history is yet to
    be understood.

    In absence of other glacial landforms, the degree of glacial modification
    of a landscape has sometimes been used as a proxy for the duration of past
    glacier cover, as for instance in geomorphological mapping \aref. However,
    cold-based glaciers have been shown to preserve landforms as fragile as
    sand beaches \aref. Yet it remains unclear how sharp the erosional
    transition to cold-based ice is, and why some glaciated regions show
    glacially preserved landscapes and others, including the Alps, do not
    \aref.

    While it has long been understood that glaciers leave a specific erosional
    imprint on the landscape, observing and quantifying the long-term erosion
    and sedimentation processes has been a challenge, not least due to the
    inaccessibility of glacier beds and the slow and stochastic nature of
    glacial erosion processes \aref. \ian{Perhaps this needs more detail?}

    More recent studies have moved away from process
    understanding towards attempts at calibrating empirical erosion laws for an entire
    glacier or set of glaciers. \citet{Koppes.etal.2015} quantified sediment
    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
    predominant control of surface air temperature on rapid glacier erosion and
    calibrating a near-square-law from glacier velocity to erosion rate.
    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
    in the outlet stream of Franz-Josef Glacier, mapped their chemical
    composition to geologic zones of distinctive glacier speed, and also
    concluded to a near-square relationship between basal sliding and glacier
    erosion, but yielding much higher erosion rates.
    \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
    38~glaciers, showing a predominant role of glacier sliding velocity over
    climate variables, yet concluding at a sub-linear relationship to erosion.

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rate and cumulative last glacial cycle erosion
    potential are analysed. Despite aggregated uncertainties on paleoclimate,
    glacier flow and glacier erosion processes, out results provide insights
    into the diversity of the Alpine glacial erosion landscape.
